---
title: "ShotQuality"
author: "Jack Weyer"
date: "5/10/2022"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

##Question 1: Find the leaders in points per game
```{r}
# Load in the data
library(tidyverse)
nba_pbp <- read.csv("nba_pbp.csv")
```

Made shot information is stored in three different columns. Let's get a single description for the play.
```{r}
library(tidymodels)

descriptions = nba_pbp %>% 
  select(descriptionPlayHome, descriptionPlayNeutral, descriptionPlayVisitor) %>% #select the three description columns
  mutate_if(is.factor, as.character) %>% #change to character
  replace(is.na(.), '') %>% #replace NAs with empty strings
  mutate(DESCRIPTION = paste0(descriptionPlayHome, descriptionPlayNeutral, descriptionPlayVisitor)) # "add" the columns together

nba_pbp$DESCRIPTION = descriptions$DESCRIPTION #add the column to df
```

Extract point value from the description of made shots. 
1) Filter to shot attempts
2) Filter to makes
3) Set Free Throws to 1, 3s to 3, and other to 2
```{r}
pointsAdded = nba_pbp %>% 
  filter(numberEventMessageType %in% c(1,3)) %>% # get made shots and free throw attempts
  filter(!str_detect(DESCRIPTION, 'MISS')) %>% #no misses.. down to point scoring
  mutate(ptVal = case_when(
    grepl('Free Throw', DESCRIPTION) ~ 1, #if free throw in description.. set to 1
    grepl('3PT', DESCRIPTION) ~ 3)) %>% # if 3pt in description.. set to 3
 mutate(ptVal = if_else(is.na(ptVal), 2, ptVal)) # else set to 2
```

Get a simple dataframe matching names with IDs
```{r}
PlayerList = nba_pbp %>% 
  select(idPlayerNBA1, namePlayer1) %>% 
  group_by(idPlayerNBA1, namePlayer1) %>% 
  na.omit() %>% 
  slice(1:1) %>% 
  arrange(namePlayer1)

# some players are never in player1.. add those in player2 but not 1
PlayerList = nba_pbp %>% 
  select(idPlayerNBA2, namePlayer2) %>% 
  group_by(idPlayerNBA2, namePlayer2) %>% 
  na.omit() %>% 
  slice(1:1) %>%
  filter(!idPlayerNBA2 %in% PlayerList$idPlayerNBA1) %>%
  rename(idPlayerNBA1 = idPlayerNBA2) %>% 
  rename(namePlayer1 = namePlayer2) %>% 
  bind_rows(PlayerList) %>% 
  arrange(namePlayer1)

# checking for players in 3 but not 1 or 2. There are none
nba_pbp %>% 
  select(idPlayerNBA3, namePlayer3) %>% 
  group_by(idPlayerNBA3, namePlayer3) %>% 
  na.omit() %>% 
  slice(1:1) %>%
  filter(!idPlayerNBA3 %in% PlayerList$idPlayerNBA1) #%>%
  #rename(idPlayerNBA1 = idPlayerNBA2) %>% 
  #rename(namePlayer1 = namePlayer2) %>% 
  #bind_rows(PlayerList) %>% 
  #arrange(namePlayer1)
```


Get Points
1) Group by Player
2) add points from each made shot
3) Join with player list
```{r}
pointsStats = pointsAdded %>% 
  group_by(idPlayerNBA1) %>% 
  summarise(totPoints = sum(ptVal)) %>% 
  arrange(desc(totPoints)) %>% 
  full_join(PlayerList) %>% # get names
  rename(PlayerID = idPlayerNBA1) %>% 
  replace(is.na(.), 0)
```

Get Games Played
1) Filter to substitutions. If a player subs in OR out of a game then they played in the game. No players played the entire game in 2021-22
2) Get players who subbed in OR out of a particular game
3) Get (distinct) players who PLAYED in a game
4) Count games played in by player, join with points table by player id
```{r}
gp = nba_pbp %>% 
  filter(numberEventMessageType==8) %>%  #substitutions. Logic is if a player subs in or out of a game, then they played in the game. This will NOT catch players who played the entire game but that doesn't happen in 2022
  select(idPlayerNBA1, idPlayerNBA2, idGame) %>% 
  group_by(idGame)

Player1s = gp %>%
  distinct(idPlayerNBA1) %>% #subbed out of a particular game 
  mutate(PlayerID = idPlayerNBA1)

Player2s = gp %>%
  distinct(idPlayerNBA2) %>% #subbed into a particular game
  mutate(PlayerID = idPlayerNBA2)

players_in_games = bind_rows(Player1s, Player2s) %>% 
  distinct(idGame, PlayerID) # count a player just once for subbing in/out

PointGameTots = players_in_games %>% 
  group_by(PlayerID) %>% 
  count() %>% # count games played in
  rename(Games = n) %>% 
  full_join(pointsStats) %>% # join with points stats
  replace_na(list(Games = 0, totPoints = 0))
```
##Solution
```{r}
PointGameTots %>% 
  group_by(namePlayer1) %>% 
  mutate(PPG = totPoints / Games) %>%
  rename(Player = namePlayer1) %>% 
  arrange(desc(PPG)) %>% 
  mutate(PPG = round(PPG,1)) %>% 
  select(Player, PPG) %>% 
  head(10)
```




##Question 2: Choose any player, which teammates does he assist the most? Who assists his shots? Include a data visualization.

Let's use Donovan Mitchell because a great deal was made about his passing (or lack thereof) to Rudy Gobert. The Jazz may also "blow it up" so it would be interesting to see how other players meshed with Mitchell.

1) Filter made shots to include Mitchell as the assister or shooter on shots with an assist
```{r}
mitchell = pointsAdded %>% # can use the pointsAdded df because assists are a subset of made shots
  filter(str_detect(DESCRIPTION, 'AST')) %>% # look for assist in the description
  filter(namePlayer2 == 'Donovan Mitchell' | 
           namePlayer1 == 'Donovan Mitchell')
```

Group by assist (from Mitchell) partners and count.
```{r}
Assists_to = mitchell %>% 
  filter(namePlayer2 == 'Donovan Mitchell') %>% 
  group_by(namePlayer1) %>% # group by assist partners
  count() %>% 
  rename(Assists = n) %>% 
  rename(Player = namePlayer1) %>%
  arrange(desc(Assists))

Assists_to %>% head(7)

Assists_to$type = 'Passing' # for use in later viz
```
Mitchell assisted to Bogdanovic the most by far. Gobert was third but only ahead of Whiteside by 4 which is significant compared to their minutes played.

Get Spida's assisted shots (Mitchell shooter)
```{r}
Assists_from = mitchell %>% 
  filter(namePlayer1 == 'Donovan Mitchell') %>% 
  group_by(namePlayer2) %>% 
  count() %>% 
  rename(Assists = n) %>% 
  rename(Player = namePlayer2) %>% 
  arrange(desc(Assists))

Assists_from %>% head(7)

Assists_from$type = 'Receiving'
```
O'Neale assisted him the most. Mitchell has much more assists than assisted shots

Create visualization
```{r}
Ast_df = bind_rows(Assists_from, Assists_to)

library(ggthemes)
  
Ast_df %>% 
  mutate(AdjAssists = if_else(type == 'Receiving', 
                              -Assists,  
                              Assists)) %>% # set assists received to negative for our diverging bar chart 
  filter(abs(AdjAssists) >=5) %>% 
  filter(!Player %in% c('Trent Forrest', 'Rudy Gay', 'Juancho Hernangomez',
                        'Eric Paschall', 'Danuel House Jr.')) %>% # get players with min 20 assist interactions
  ggplot(aes(x=AdjAssists, y=Player, fill=type))+
   scale_y_discrete(limits=c("Joe Ingles","Hassan Whiteside","Rudy Gobert","Jordan Clarkson","Mike Conley","Royce O'Neale", 'Bojan Bogdanovic'))+ #set the order
  geom_bar(stat = 'identity')+
  scale_fill_manual(values = c('#002B5C','#F9A01B'))+ #jazz colors
  theme_fivethirtyeight()+ #538 total interactions -> 538 theme
  labs(y = '', x = 'Assists', title = "Donovan Mitchell's Assists (navy) and Assisted shots (gold)",
       subtitle = "2021-22 season (min. 20 total assists)",
       fill = "Assist type") +
  theme(plot.title = element_text(hjust = .9, size = 15))
```

##Question 3: Which first names play the most minutes? Last names? In addition to those two, find your own interesting name-related statistic or leaderboard.

Plan: 
Get minutes played by player
Split on First space
Split on next space
Aggregate

Load in quarter starters
Filter OG pbp data to substitutions only & select desired columns
```{r}
qtrs = read.csv("QuarterStarters.csv")

AllSubs = nba_pbp %>% 
  filter(numberEventMessageType==8) %>% 
  arrange(idGame) %>% 
 # group_by(idGame) %>%
  select(idGame, numberEvent, numberPeriod, idPlayerNBA1, idPlayerNBA2, minuteRemainingQuarter, secondsRemainingQuarter)
```

1) Set dummy variables for quarter and overtime STARTS for each sub
2) Group by player and count up their quarter and overtime STARTS
3) Count up the seconds from these periods, ASSUMING THEY PLAY THE ENTIRE PERIOD
```{r}
startingQtrSeconds = qtrs %>% 
  mutate(QTR_dummy = if_else(PERIOD < 5, 1, 0)) %>% 
  mutate(OT_dummy = if_else(PERIOD > 4,1,0)) %>% 
  group_by(PLAYER_ID) %>% 
  summarise(totQtrs = sum(QTR_dummy),
            totOTs = sum(OT_dummy),
            seconds = totQtrs*12*60 + totOTs*5*60) %>% 
  rename(idPlayerNBA1 = PLAYER_ID)
```

Add to player list
```{r}
PlayerList = full_join(PlayerList, startingQtrSeconds, by="idPlayerNBA1") %>% 
  select(-totQtrs, -totOTs) %>% 
   replace(is.na(.), 0) # this assumes players play the whole period they start in. Need to subtract/add from this number using the subs
```

1) if subbed in ... add time remaining in period to their cumulative seconds played in season
2) if subbed out ... subtract time remaining in period from their cumulative seconds played in season
3) get minutes played by player in season
```{r}
AllSubs = AllSubs %>% 
  mutate(totSecondsLeft = minuteRemainingQuarter*60 + secondsRemainingQuarter) %>% # get seconds left in period
  rename(sub_out = idPlayerNBA1) %>% 
  rename(sub_in = idPlayerNBA2)

addTime = AllSubs %>% 
  group_by(sub_in) %>% 
  summarise(added_time = sum(totSecondsLeft)) %>% 
  replace(is.na(.), 0) %>% 
  rename(idPlayerNBA1 = sub_in)

subtractTime = AllSubs %>% 
   group_by(sub_out) %>% 
    summarise(subtract_time = -sum(totSecondsLeft)) %>% 
    replace(is.na(.), 0) %>% 
    rename(idPlayerNBA1 = sub_out)

PlayerList = PlayerList %>% full_join(addTime)
PlayerList = PlayerList %>% full_join(subtractTime)

PlayerList$added_time = PlayerList$added_time %>% 
  replace(is.na(.), 0)

PlayerList$subtract_time = PlayerList$subtract_time %>% 
  replace(is.na(.), 0)


totalMinutes = PlayerList %>%
  #replace(is.na(PlayerList$added_time), 0) %>% 
  mutate(totSeconds = seconds + added_time + subtract_time) %>% #subtract time is already negative
  mutate(minutes = totSeconds / 60) %>% # get minutes
  arrange(desc(minutes)) %>% 
  select(idPlayerNBA1, namePlayer1, minutes)
```

Get minutes played by first name
```{r}
totalMinutes$namePlayer1 = as.character(totalMinutes$namePlayer1)

totalMinutes = totalMinutes %>% 
  mutate(first_name = strsplit(namePlayer1, " +")[[1]][1]) %>% #first name variable 
  mutate(last_name = strsplit(namePlayer1, " +")[[1]][2]) #last name variable

totalMinutes %>% 
  group_by(first_name) %>% 
  summarise(mins = round(sum(minutes),0)) %>% 
  arrange(desc(mins)) %>% 
  head(10)
```
Josh, Kevin, and Jalen are the winners.

```{r}
totalMinutes %>% 
  group_by(last_name) %>% 
  summarise(mins = round(sum(minutes),0)) %>% 
  arrange(desc(mins)) %>% 
  head(10)
```
Green, Williams, and Johnson are the winners. Josh Green wins the average player award

My visualization.. first initial
```{r}
totalMinutes %>% 
  mutate(first_letter = substr(first_name,1,1)) %>% 
  group_by(first_letter) %>% 
  summarise(mins = sum(minutes)) %>% 
  filter(mins>0) %>% 
  ggplot(aes(x=first_letter, y=mins, fill = 'red'))+
  geom_bar(stat = 'identity')+
  labs(x='',
       y = "Minutes Played",
       title = "Which first initial played the most minutes?",
       subtitle = "2021-22 NBA season")+
  theme_fivethirtyeight()+
  theme(legend.position = "none")
```



